Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-07-01 paint-contour-tree.tex
Many reviewer comments handled

\begin{abstract}
We revisit the classical problem of computing the \emph{contour tree}
of a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
The contour tree is a fundamental topological data structure that tracks
the evolution of level sets of $f$ and 
has numerous applications in data analysis and visualization.

Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
but these can be $\Omega(n)$). 
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first input-sensitive algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
Our results match all existing upper bounds, but also
characterize the amount of sorting required to compute the contour tree.
This leads to an almost linear (inverse Ackermann overhead) time algorithm for contour tree computation for many
classes of natural instances. 
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree, 
allowing us to complement our algorithmic results with matching lower bounds for contour tree computation.

@@

Geometric data (potentially massive is size) is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$. 
For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e. a contour.

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold
%  or alternately consider
% $\MM$ in $\RR^{d+1}$
with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. More generally, these are the events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
An edge connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

@@

Suppose we have $f: \MM \mapsto \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{definition}
\label{def:path} 
For a contour tree $T$, a \emph{leaf path} is any path in $T$ containing a leaf, 
which is also monotonic in the height values of its vertices. 
Then a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
\end{definition}

\begin{theorem} 
\label{thm:main-alg} 
Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).

@@

\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e., $d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
%
%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
requires this time in the worst case, and our algorithm matches this bound.

\paragraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
characterize hard instances.
% explain \emph{when}
% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
%It is intriguing
%that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
This is tangentially related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. In general, our analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.

There are numerous algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
This result forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and in its contour tree $\cC(\MM)$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced4}
    \hspace{.3in}
    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
    \caption{On left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
    On right (from left to right), the contour, join and split trees.}
    \label{fig:sorting}
\end{figure}

@@

We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear (PL) function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
and all other valued are obtained by interpolation.

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time traversals of neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$.

@@

\begin{definition} \label{def:deg} The Morse up-degree of $x$ is the number of $(f(x) + \delta)$-contours in $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The Morse down-degree is the number of $(f(x) - \delta)$-contours in $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.
\end{definition}

@@

\begin{definition} \label{def:points} A point $x$ is categorized as follows.
\begin{asparaitem}
	\item Regular: Both Morse up-degree and down-degrees are $1$.
	\item Maximum: Morse up-degree is $0$.
	\item Minimum: Morse down-degree is $0$.
	\item Morse Join: Morse up-degree is strictly greater than $1$.
	\item Morse Split: Morse down-degree is strictly greater than $1$.
\end{asparaitem}
Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
Maxima and minima are \emph{extrema}.
\end{definition}

The set of critical points is denoted by $\cV(f)$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maxima or minima.

\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

@@

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV= \cV(f)$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
\end{definition}

@@

\subsection{Some technical remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have $>1$ degree in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explictly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

\section{A tour of the new contour tree algorithm} \label{sec:approach}

Our final algorithm is quite technical and has numerous moving parts. This section
provides a high level view of the entire result, and highlights the main ideas
and intuition. 
\SubmitVer{The full version with all the details can be found in the appendix.}
It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
terrain. In the interest of presentation, the definitions and theorem statements in this section will
slightly differ from those in the main body. They may also differ from the original definitions proposed
in earlier work.

\paragraph{Do not globally sort:} Arguably, the starting point for this work is \Fig{relative}. We have two terrains
with exactly the same contour tree, but different orderings of (heights of) the critical points.
Turning it around, we cannot deduce the full height ordering of critical points from the contour tree. Therefore, clearly
sorting all critical points is computationally unnecessary for constructing the contour tree.
In \Fig{sorting}, the contour tree basically consists of two balanced binary trees, one of the joins,
another of the splits. Again, it is not necessary to know the relative ordering between the mounds
on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
information is necessary: on the left, the little valleys are higher than the big central valley,
and this is reflected in the contour tree. More generally, leafs paths in the contour tree have points
in sorted order, but incomparable points in the tree are unconstrained.
How do we sort exactly what is required, without knowing the contour tree in advance?

@@

\begin{definition} 
\label{def:int-criticalJoin}
The critical join tree $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

(Abusing notation, we just call this the join tree.)
Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. They simply form a long path. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e. unlike \Fig{sorting}) 
the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
to get the contour tree requires having the split vertices in the join tree. Without
this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.

@@

\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there only exists a single minima.
(Our real definition is more complicated, and involves simplicial complexes
that allow additional ``trivial'' minima.)
In such a complex, there exists a non-ascending path from any point to this unique minima.
Analogously, we can define maximum dominant simplicial complexes,
and both are collectively called extremum dominant.

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
The best way to think of our procedure is a meteorology analogy. Take an arbitrary maxima $x$,
and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
is a set of contours, given by the ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we can recur this procedure on them.

>>>>

\begin{abstract}
We revisit the classical problem of computing the \emph{contour tree}
of a scalar field $f:\MM \to \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
The contour tree is a fundamental topological structure that tracks
the evolution of level sets of $f$ and 
has numerous applications in data analysis and visualization.

Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
but these can be $\Omega(n)$). 
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
Our results match all existing upper bounds, but also
characterize the amount of sorting required to compute the contour tree.
This leads to an almost linear (inverse Ackermann overhead) time algorithm for contour tree computation for many
classes of natural instances. 
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree, 
allowing us to complement our algorithmic results with matching lower bounds for contour tree computation.

@@

Geometric data (potentially massive in size) is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological structure used to represent and visualize
the function $f$. 
For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour.

It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
%  or alternately consider
% $\MM$ in $\RR^{d+1}$
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. More generally, these are the events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
These points are the ``events" where the contour topology changes and which form the vertices of the contour tree. 
An edge of the contour tree connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

@@

Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{definition}
\label{def:path} 
For a contour tree $T$, a \emph{leaf path} is any path in $T$ containing a leaf, 
which is also monotonic in the height values of its vertices. 
Then a \emph{path decomposition}, $P(T)$, is a partition of the vertices of $T$ into a set of vertex disjoint leaf paths.  
\end{definition}

\begin{theorem} 
\label{thm:main-alg} 
Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \to \RR$, where the contour tree $T$
has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).

@@

\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e.\ $d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
%
%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

In other words, for any path decomposition, $P$, we can design a hard family where any algorithm
requires $\sum_{p \in P} |p|\log |p|$ time in the worst case, and our algorithm matches this bound.

\paragraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
characterize hard instances.
% explain \emph{when}
% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
%It is intriguing
%that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. In general, our analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.

There are numerous algorithms for computing contour trees.
An $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, was presented by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
This result forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
in $\MM$ and in its contour tree $\cC(\MM)$.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced4}
    \hspace{.3in}
    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
    \caption{On left, a surface with a balanced contour tree, but whose join and split trees have long tails.  
    On right (from left to right), the contour, join and split trees.}
    \label{fig:sorting}
\end{figure}

@@

We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear (PL) function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
and all other values are obtained by linear interpolation.

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
we allow for other boundary facets, to capture the resulting surface pieces after our algorithm make a horizontal cut.

We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time traversals of neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$, 
where $f(x)$ is the ``height" of a point $x \in \MM$, which is encoded in the representation of $\MM$. 
Specifically, rather than writing our input as $(\MM,f)$, we abuse notation and typically just write $\MM$ to denote the lifted complex.

@@

\begin{definition} \label{def:deg} The \emph{Morse up-degree} of $x$ is the number of $(f(x) + \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$. The \emph{Morse down-degree} is the number of $(f(x) - \delta)$-contours of $f|B_\eps(x)$
as $\delta, \eps \rightarrow 0^+$.
\end{definition}

@@

\begin{definition} \label{def:points} A point $x$ is categorized as follows.
\begin{asparaitem}
	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
	\item \emph{Maximum:} Morse up-degree is $0$.
	\item \emph{Minimum:} Morse down-degree is $0$.
	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
\end{asparaitem}
Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
Maxima and minima are \emph{extrema}.
\end{definition}

The set of critical points is denoted by $\cV(f)$.
Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
A contour is called \emph{critical}, if it contains a critical point, and it is called \emph{regular} otherwise.

The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maximum or minimum.

\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are \emph{equivalent} 
if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
such that no $x \in p$ belongs to a critical contour.
\end{definition}

@@

\begin{definition} \label{def:tree} The \emph{contour tree} is the graph on vertex set $\cV= \cV(f)$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
\end{definition}

@@

\subsection{Some technical remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

\section{A tour of the new contour tree algorithm} \label{sec:approach}
Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
can be explained clearly.  
\SubmitVer{Therefore, in the remaining few allotted pages, 
rather than providing an incomplete subset of the technical content, we provide a high level view of the entire result, 
which we view as essential for understanding our approach. The technical details can be found in the full version of the paper \cite{rs-mvrms-14}.}
\FullVer{Therefore, here we first provide a high level view of the entire result, 
which we view as essential for understanding our approach. }

In the interest of presentation, the definitions and theorem statements in this section will
slightly differ from those in the main body. They may also differ from the original definitions proposed
in earlier work.

% Our final algorithm is quite technical and has numerous moving parts. This section
% provides a high level view of the entire result, and highlights the main ideas
% and intuition. 
% \SubmitVer{The full version with all the details can be found in the appendix.}
% It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
% terrain. In the interest of presentation, the definitions and theorem statements in this section will
% slightly differ from those in the main body. They may also differ from the original definitions proposed
% in earlier work.

\paragraph{Do not globally sort:} Arguably, the starting point for this work is \Fig{relative}. We have two terrains
with exactly the same contour tree, but different orderings of (heights of) the critical points.
Turning it around, we cannot deduce the full height ordering of critical points from the contour tree. Therefore, clearly
sorting all critical points is computationally unnecessary for constructing the contour tree.
In \Fig{sorting}, the contour tree basically consists of two balanced binary trees, one of the joins,
another of the splits. Again, it is not necessary to know the relative ordering between the mounds
on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
information is necessary: on the left, the little valleys are higher than the big central valley,
and this is reflected in the contour tree. More generally, leaf paths in the contour tree have points
in sorted order, but incomparable points in the tree are unconstrained.
How do we sort exactly what is required, without knowing the contour tree in advance?

@@

\begin{definition} 
\label{def:int-criticalJoin}
The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
and $v$ is connected to this component (in $\MM$). 
\end{definition}

Abusing notation, for now we will just call this the join tree. (In Section~\ref{sec:extreme} we carefully handle the distinction 
between the join tree and the critical join tree.)
Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. They simply form a long path. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
to get the contour tree requires having the split vertices in the join tree. Without
this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.

@@

\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
(Our real definition is more complicated, and involves simplicial complexes
that allow additional ``trivial'' minima.)
In such a complex, there exists a non-ascending path from any point to this unique minimum.
Analogously, we can define maximum dominant simplicial complexes,
and both are collectively called extremum dominant.

We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
The best way to think of our procedure is a meteorology analogy. Take an arbitrary maximum $x$,
and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
while there may be numerous disconnected dry parts. The interface between the dry and wet parts
is a set of contours, given by the ``water line". The wet part is clearly maximum dominant, since all
wet points have a non-descending path to $x$. So we can simply cut along the interface
contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
we are left with a set of disconnected dry parts, and we can recur this procedure on them.
